\documentclass{article}
\usepackage[utf8]{inputenc}

\title{Lab 10: Linear Regression\\STA 360/602}
\author{Rebecca C. Steorts}

\usepackage{natbib}
\usepackage{graphicx}
\usepackage{float}
\usepackage{amsmath}
\usepackage{verbatim}
\usepackage{bbm}



\begin{document}

\maketitle

\section*{Introduction}

The topic of this lab is linear regression. It follows exercise 9.1 in Hoff very closely. Open the data file with the following code: 
\begin{verbatim}
swim <- read.table("~/.../swim.dat", sep=' ')
\end{verbatim}
replacing the ellipses with the appropriate directory name. The file contains times (in seconds) of four high school swimmers swimming 50 yards. There are 6 times for each student, taken every two weeks. Each row corresponds to a swimmer and a higher column index indicates a later date. 

\section*{Task 1} We will fit a separate linear regression model for each swimmer, with swimming time as the response and week as the explanatory variable. Let $Y_{i}\in \mathbbm{R}^{6}$ be the 6 recorded times for swimmer $i.$ Let
\[X_i =
\begin{bmatrix}
    1 & 1  \\
    1 & 3 \\ 
    ... \\
    1 & 9\\
    1 & 11
\end{bmatrix}
\] be the design matrix for swimmer $i.$ Then we use the following linear regression model: 
\begin{align*}
    Y_i &\sim \mathcal{N}_6\left(X\beta_i, \tau_i^{-1}\mathcal{I}_6\right) \\
    \beta_i &\sim \mathcal{N}_2\left(\beta_0, \Sigma_0\right) \\
    \tau_i &\sim \text{Gamma}(a,b).
\end{align*}
Derive full conditionals for $\beta_i$ and $\tau_i.$

\section*{Task 2}

Complete the prior specification by choosing $a,b,\beta_0,$ and $\Sigma_0.$ Let your choices be informed by the fact that times for this age group tend to be between 22 and 24 seconds. 

\section*{Task 3}

Code a Gibbs sampler to fit each of the models. For each swimmer $i,$ obtain draws from the posterior predictive distribution for $y_i^{*},$ the time of swimmer $i$ if they were to swim two weeks from the last recorded time.

\section*{Task 4}

The coach has to decide which swimmer should compete in a meet two weeks from the last recorded time. Using the posterior predictive distributions, estimate $\text{Pr}\left\{y_i^{*}=\text{max}\left(y_1^{*},y_2^{*},y_3^{*},y_4^{*}\right)\right\}$ for each swimmer $i$ and use these probabilities to make a recommendation to the coach. 

\end{document}
