
---
title: "Module 7: Introduction to Gibbs Sampling"
author: "Rebecca C. Steorts"
output: 
     beamer_presentation:
      includes: 
          in_header: custom2.tex
font-size: 8px
---

Agenda
===

In this lab, we will deriving conditional distributions, code a Gibbs sampler, and analyze the output of the Gibbs sampler. 

Problem Statement
===

Consider the following Exponential model for observation(s) $\bf{x}=(x_1,\ldots,x_n).$\footnote{The data can be found in \text{data-exponential.csv}.}:
$$ p(x|a,b) = a b \exp(- a b x) I(x>0),$$ where the $x_i$ are assumed to be iid for $i=1,\ldots n.$
and suppose the prior is 
$$ p(a,b) = \exp(- a - b)I(a,b>0). $$
You want to sample from the posterior $p(a,b|x_{1:n})$.  You may assume that $$a=0.25, b=0.25$$ when coding up your Gibbs sampler.

Tasks
===


1. Find the conditional distributions needed for implementing a Gibbs sampler.
2. Code up your own Gibbs sampler using part (1). 
3. Run the Gibbs sampler, providing convergence diagnostics. 
4.  Plot a histogram or a density estimate of the estimated posterior using (2) and (3).
5.  How do you know that your estimated posterior in (3) is reliable? 


Task 1:
===
Consider the following Exponential model for observation(s) $x=(x_1,\ldots,x_n).$\footnote{Please note that in the attached data there are 40 observations, which can be found in \text{data-exponential.csv}.}:
$$ p(x|a,b) = a b \exp(- a b x) I(x>0)$$
and suppose the prior is 
$$ p(a,b) = \exp(- a - b)I(a,b>0). $$
You want to sample from the posterior $p(a,b|x)$. 

Task 1: Conditional distributions
===
\begin{align*}
p(\boldsymbol{x}|a,b) &= \prod_{i=1}^n p(x_i|a,b) \\
&= \prod_{i=1}^n ab\exp(-abx_i) \\
&= (ab)^n\exp\left(-ab\sum_{i=1}^nx_i\right).
\end{align*}
The function is symmetric for $a$ and $b$, so we only need to derive $p(a|\boldsymbol{x},b)$.  

Task 1: Conditional distributions
===
This conditional distribution satisfies
\begin{align*}
p(a|\boldsymbol{x},b) &\propto_a p(a,b,\boldsymbol{x}) \\
&= p(\boldsymbol{x}|a,b)p(a,b) \\
&= \textcolor{blue}{\text{fill in full details for lab this week}}
\end{align*}

Task 2: Gibbs sampling code
===
```{r}
knitr::opts_chunk$set(cache=TRUE)
library(MASS)
data <- read.csv("data-exponential.csv", header = FALSE)
```

Task 2: Gibbs sampling code
===
```{r}
#######################################
# This function is a Gibbs sampler
# 
# Args
#   start.a: initial value for a
#   start.b: initial value for b
#   n.sims: number of iterations to run
#   data:  observed data, should be in a 
            # data frame with one column
#
# Returns:
#   A two column matrix with samples 
     #   for a in first column and
# samples for b in second column
#######################################
```

Task 2: Gibbs sampling code
===
```{r}
sampleGibbs <- function(start.a, start.b, n.sims, data){
  # get sum, which is sufficient statistic
  x <- sum(data)
  # get n
  n <- nrow(data)
  # create empty matrix, allocate memory for efficiency
  res <- matrix(NA, nrow = n.sims, ncol = 2)
  res[1,] <- c(start.a,start.b)
  for (i in 2:n.sims){
    # sample the values
    res[i,1] <- rgamma(1, shape = n+1, 
                       rate = res[i-1,2]*x+1)
    res[i,2] <- rgamma(1, shape = n+1, 
                       rate = res[i,1]*x+1)
  }
  return(res)
}
```

Task 3: Run the Gibbs sampler 
===
```{r}
# run Gibbs sampler
n.sims <- 10000
# return the result (res)
res <- sampleGibbs(.25,.25,n.sims,data)
head(res)
```



Task 4
===

Plot a histogram or a density estimate of the estimated posterior using tasks (2) and (3).

Finish this for homework. 


Task 5
=== 

How do you know that your estimated posterior in task (3) is reliable? 

Finish for homework.