
---
title: "Lab 3: Intro to Decision Theory"
author: "Xingyu Yan and Rebecca C. Steorts"
output: 
     pdf_document:
      includes: 
font-size: 8px
---

# Task 1

```{r, cache=TRUE}
# set seed 
set.seed(123)

# data
sum_x = 1
n = 30
# prior parameters
a = 0.05; b = 1
# posterior parameters
an = a + sum_x
bn = b + n - sum_x
th = seq(0,1,length.out = 100)
like = dbeta(th, sum_x+1,n-sum_x+1)
prior = dbeta(th,a,b)
post = dbeta(th,sum_x+a,n-sum_x+b)
```

We now consider the loss function. 
```{r,cache=TRUE}
# compute the loss given theta and c 
loss_function = function(theta, c){
  if (c < theta){
    return(10*abs(theta - c))
  } else{
    return(l = abs(theta - c))
  }
}
```

We now write a function **posterior_risk** which is a function of c, parameters a_prior and b_prior for the prior distribution of $\theta$, the summation of $x_i$ sum_x, the number of observations n, and also the number of random draws s. 
```{r}
# compute the posterior risk given c 
# s is the number of random draws 
posterior_risk = function(c, a_prior, b_prior, sum_x, n, s = 30000){
  # randow draws from beta distribution 
  a_post = a_prior + sum_x
  b_post = b_prior + n - sum_x
  theta = rbeta(s, a_post, b_post)
  loss <- apply(as.matrix(theta),1,loss_function,c)
  # average values from the loss function
  risk = mean(loss)
}
# a sequence of c in [0, 0.5]
c = seq(0, 0.5, by = 0.01)
post_risk <- apply(as.matrix(c),1,posterior_risk, a, b, sum_x, n)
head(post_risk)
```

We then look at the Posterior expected loss (posterior risk) for disease prevelance versus c. 
```{r, cache=TRUE}
# plot posterior risk against c 
plot(c, post_risk, type = 'l', col='blue', 
    lwd = 3, ylab ='p(c, x)' )

# minimum of posterior risk occurs at c = 0.08
(c[which.min(post_risk)])
```

# Task 2

We now consider task 2. We set $a  = 0.05, 1, 0.05$ and $b = 1, 2, 10$. If we have different prior, the posterior risk is minimized at different c values. The optimal c depends on not only the data, but also the prior setting.  
```{r,cache=TRUE}
# set prior
as = c(0.05, 1, 0.05); bs = c(1, 1, 10)
post_risk = matrix(NA, 3, length(c))

# for each pair of a and b, compute the posterior risks
for (i in 1:3){
  a_prior = as[i]
  b_prior = bs[i]
  
  post_risk[i,] = apply(as.matrix(c), 1, posterior_risk, a_prior, b_prior, sum_x, n)
}

plot(c, post_risk[1,], type = 'l', col='blue', lty = 1, yaxt = "n", ylab = "p(c, x)")
par(new = T)
plot(c, post_risk[2,], type = 'l', col='red', lty = 2, yaxt = "n", ylab = "")
par(new = T)
plot(c, post_risk[3,], type = 'l', lty = 3, yaxt = "n", ylab = "")
legend("bottomright", lty = c(1,2,3), col = c("blue", "red", "black"), 
       legend = c("a = 0.05 b = 1", "a = 1 b = 1", "a = 0.05 b = 5"))
```

Note there is a more automated solution but this is the most simple one and is completely correct. 



