
---
title: "Module 9: Logistic Regression"
author: "Rebecca C. Steorts"
output: 
     beamer_presentation:
      includes: 
          in_header: custom2.tex
font-size: 8px
---


Agenda
===

- 1986 Challenger explosion
- What happened?
- Background
- Logistic regression

The Challenger Case Study
===

On 28 January 1986, the Space Shuttle Challenger broke apart, 73 seconds into
flight. All seven crew members died. The cause of the disaster was the failure of
an o-ring on the right solid rocket booster.

O-rings
===

- O-rings help seal the joints of different segments of the solid rocket boosters. 

- We learned after this fatal mission that o-rings can fail at extremely low temperature. 



Motivations and goals
===

- In 1986, the Challenger space shuttle exploded as it took off.
- The question of interest was what happened and could it have been prevented?
- We will revisit not just the challenger data, but other missions to understand the relationship between o-ring failure and temperature. 
- To understand this, we need to learn about logistic regression. 


Loading the Faraway Package
===
```{r}
library(faraway)
data("orings")
orings[1,] <- c(53,1)
head(orings)
```

Space Shuttle Missions
===

The 1986 crash of the space shuttle Challenger was linked to failure of o-ring seals in the rocket engines. 

Data was collected on the 23 previous shuttle missions, where the following variables were collected:

- temperate for each mission
- damage to the number of o-rings (out of a total of six)

Plot
===
```{r}
plot(damage~temp, data=orings, xlab="temperature (F)",
     ylab="damage (out of 6)",
     pch=16, col="red", ylim=c(0,6))
```



Plot
===
```{r}
library(ggplot2)
geom_boxplot(outlier.colour="black", outlier.shape=14,
             outlier.size=2, notch=FALSE)
damage <- as.factor(orings$damage)
temp <- orings$temp
head(damage) 
```

Boxplot of temperature versus o-ring failure 
===
```{r, warning=FALSE ,echo=FALSE}
p <- ggplot(orings, aes(x=as.factor(orings$damage), y=temp)) + 
  geom_boxplot()
p + ggtitle("") +
  xlab("Damage") + ylab("Temp (F)")
```

Response and covariate
===

- The response is the damage to the o-ring (in each shuttle launch). 

- The covariate is the temperature (F) in each shuttle launch. 

Notation and Setup
===

- Let $p_i$ be the probability that o-ring $i$ fails. 

- The corresponding **odds of failure** are $$\frac{p_i}{1-p_i}.$$

Notation and Setup
===

- The probability of failure $p_i$ is between $[0,1]$

- The odds of failure is any real number. 

Logistic Regression
===

The response 

\begin{align}
Y_i \mid p_i &\sim \text{Bernoulli}(p_i)
\end{align}
for $i=1,\ldots,n.$

The logistic regression model writes that the logit of the probability  $p_i$
is a linear function of the predictor variable(s) $x_i$: 

\begin{align}
\text{logit}(p_i)  &:= \log(\frac{p_i}{1-p_i}) = \beta_0 + \beta_1x_i.
\end{align}

Interpretation of Co-efficients
===

- The regression co-efficients $\beta_0$,  $\beta_1$ are directly related to the log odds $log(\frac{p_i}{1-p_i})$
and not $p_i.$

- For example, the intercept $\beta_0$ is the $log(\frac{p_i}{1-p_i})$ for observation $i$ when the predictor takes a value of 0.

- The slope $\beta_1$ refers to the change in the expected log odds of failure of an o-ring for a decrease in temperature. 

Intuition of Model
===

We assume our 23 data points are **conditionally independent**. 

$$\text{Pr(failure = 1)} = \frac{\exp\{\beta_0 + \beta_1 \times \text{temp}\}}{1+ \exp\{\beta_0 + \beta_1 \times \text{temp}\}}$$
\begin{align}
&\text{failure}_1,\ldots, \text{failure}_{23} \mid \beta_0, \beta_1, \text{temp}_1,\ldots, \text{temp}_{23} \\
& \sim 
\prod_i 
\left(\frac{\exp\{\beta_0 + \beta_1 \times \text{temp}_i\}}{1+ \exp\{\beta_0 + \beta_1 \times \text{temp}_i\}}\right)^{\text{failure}_i} \\
&\times
\left(\frac{1}{1+ \exp\{\beta_0 + \beta_1 \times \text{temp}_i\}}\right)^{\text{1-failure}_i}
\end{align}



Exercise
===

Assume that $\log(\frac{p_i}{1-p_i}) = \beta_0 + \beta_1x_i.$

Show that $$p_i = \frac{e^{\beta_0 + \beta_1x_i}}{e^{\beta_0 + \beta_1x_i} + 1}.$$

This shows that logit function guarantees that the probability $p_i$ lives in $[0,1].$



Bayesian Logistic Regression
===
Recall that 
\begin{align}
Y_i \mid p_i &\sim \text{Bernoulli}(p_i)
\end{align}
for $i=1,\ldots,n.$

\begin{align}
\text{logit}(p_i)  &:= \log(\frac{p_i}{1-p_i}) = \beta_0 + \beta_1x_i.
\end{align}

\textbf{How can we build minimal Bayesian prior knowledge?}


Priors on $\beta_0$ and $\beta_1$
===

Conjugate priors do not exists on $\beta_0$ and $\beta_1.$

We will consider the following weakly informative priors:

\begin{align}
\beta_0 &\sim \text{Normal}(0,1000) \\
\beta_1 &\sim \text{Normal}(0,1000) \\
\end{align}

Posterior sampling
===

Since we cannot find the posterior in closed form, we will resort to MCMC to approximate inference regarding 
$\beta_0, \beta_1.$

We can do this easily using the \texttt{logitMCMC} function in the \texttt{MCMCpack} R package. 

This package implements a random walk Metroplis algorithm. 

Posterior sampling
===
```{r}
library(MCMCpack)
failure <- orings$damage
temperature <- orings$temp
output <- MCMClogit(failure~temperature, 
                    mcmc=1000, b0=0, B0=0.001)
```

Traceplots
===
\footnotesize
```{r}
plot(output)
```

Summary
===
\footnotesize
```{r}
summary(output)
```

Simulating Posterior Prediction
===

Given a certain temperature, we can simulate the results of future space shuttle launches using the posterior predictive distribution. 

Suppose that on launch day, it's 80 degrees (F). 

How would we simulate a predictive probability that a o-ring would fail? 


Simulating Posterior Prediction
===

```{r}
library(boot)
temp <- 80
fail.prob <- inv.logit(output[,1]+ temp*output[,2])
y.pred <- rbinom(2100, size=1, prob=fail.prob)
```


Simulating Posterior Prediction
===

```{r}
barplot(table(y.pred))
```

Your Turn
===
  
Suppose that it's very cold, 20 F.  

How would we simulate a predictive probability that a o-ring would fail? 

- What does your group think intuitively?
- Code up a simulation of a posterior prediction and what do you find? 

<!-- Your Turn -->
<!-- === -->

<!-- ```{r} -->
<!-- library(boot) -->
<!-- temp <- 20 -->
<!-- fail.prob <- inv.logit(output[,1]+ temp*output[,2]) -->
<!-- y.pred <- rbinom(2100, size=1, prob=fail.prob) -->
<!-- ``` -->

<!-- Your Turn -->
<!-- === -->

<!-- ```{r} -->
<!-- barplot(table(y.pred)) -->
<!-- ``` -->

Summary
===

- Case Study on Challenger explosion
- Linear relationship between odds or log odds of failure and temperature seems reasonable implies use logistic regression
- What does logistic regression look like? 
- We use the Metropolis algorithm via R
- What did we learn from the case study? 
- 
