---
title: "Lab 5: Rejection sampling"
author: "Olivier Binette"
date: "28/08/2020"
fontsize: 11pt
output: 
  beamer_presentation:
    include:
      in_header: preamble.tex
---

# Agenda

1. Rejection sampling
2. Lab 5 Tasks 1-3
3. Questions / Office Hours

# 
\section{Rejection sampling}

# Rejection sampling

**The problem:** You want to sample from a distribution on $\mathbb{R}$ given either:\pause

1. its probability density function $p(\theta)$; or\pause

2. some function $f(\theta) \propto p(\theta)$.\pause

**Example:**

- In a Bayesian framework, we want to sample from the posterior distribution $p(\theta \mid x) \propto p(x \mid \theta) p(\theta) = f(\theta)$.
\pause

- You might want call `rnorm` and `rgamma`, use a parametric bootstrap, approximate the $p$-value corresponding to a complex null hypothesis, etc.

# Rejection sampling

**Summary:**

- Sampling from probability distributions is a fundamental problem in statistics and computer science.
\pause

- Often you only have access to the probability density function or something proportional to it.
\pause

- Rejection sampling is *one* way to address this.

# Rejection sampling

The area under the graph of a function $f$ is the set of points $(x,y)$ such that $0 \leq y \leq f(x)$.\pause

**Fundamental lemma of rejection sampling:** Let $f$ be a positive and integrable function. If $(X,Y)$ is uniformly distributed under the graph of $f$, then the marginal probability density function of $X$ is proportional to $f$.

# Rejection sampling

\begin{center}
  \includegraphics[width=\linewidth]{figures/fig1}
\end{center}

# Rejection sampling

\begin{center}
  \includegraphics[width=\linewidth]{figures/fig2}
\end{center}

# Rejection sampling

\begin{center}
  \includegraphics[width=\linewidth]{figures/fig3}
\end{center}

# Rejection sampling

\begin{center}
  \includegraphics[width=\linewidth]{figures/fig4}
\end{center}

# Rejection sampling

\begin{center}
  \includegraphics[width=\linewidth]{figures/fig5}
\end{center}

# Rejection sampling

\begin{center}
  \includegraphics[width=\linewidth]{figures/fig6}
\end{center}

# Rejection sampling

\begin{center}
  \includegraphics[width=\linewidth]{figures/fig7}
\end{center}

# Rejection sampling

\begin{center}
  \includegraphics[width=\linewidth]{figures/fig8}
\end{center}

# Rejection sampling

\begin{center}
  \includegraphics[width=\linewidth]{figures/fig9}
\end{center}

# Rejection sampling

\begin{center}
  \includegraphics[width=\linewidth]{figures/fig10}
\end{center}

# Rejection sampling

\begin{center}
  \includegraphics[width=\linewidth]{figures/fig11}
\end{center}

# Rejection sampling

\begin{center}
  \includegraphics[width=\linewidth]{figures/fig12}
\end{center}

# Rejection sampling

\begin{center}
  \includegraphics[width=\linewidth]{figures/fig13}
\end{center}

# Rejection sampling

\begin{center}
  \includegraphics[width=\linewidth]{figures/fig14}
\end{center}

# Rejection sampling

\begin{center}
  \includegraphics[width=\linewidth]{figures/fig15}
\end{center}

# Rejection sampling

\begin{center}
  \includegraphics[width=\linewidth]{figures/fig16}
\end{center}

# Rejection sampling

\begin{center}
  \includegraphics[width=\linewidth]{figures/fig17}
\end{center}

# Rejection sampling

## Rejection sampling Algorithm

**Input:**

- An integrable function $f \geq 0$ and an enveloppe $g \geq f$ which you can sample from.\footnote{You can sample from the density proportional to $g$.}
\pause

**Output:**

- A sample $X$ distributed following the density proportional to $f$.
\pause

**Algorithm:**

1. Sample $X \sim g$ and $Y \mid X \sim \text{unif}(0, g(X))$.
\pause

2. If $Y < f(X)$, then return $X$. Otherwise go back to step 1.

# Rejection sampling

**Note:**

- The function $g$ is called the *enveloppe* function, and the corresponding distribution is the *proposal* distribution, or the *instrumental* distribution.

- The function $f$ is called the *target*.

#

\section{Lab 5}

# Lab 5

We want to sample from the density proportional to
$$
  f(x) = \sin^2(\pi x),\quad x \in [0,1],
$$
using rejection sampling.
\pause

We'll consider two proposal distributions: 

- $Unif(0,1)$
- $\text{Beta}(2,2)$


# Lab 5

**Task 1:** Plot $f(x)$ and the Unif(0,1) density. Sample from $f(x)$ using the Unif(0,1) pdf as an enveloping function.

```{r, echo=FALSE, include=FALSE}
source(url("https://gist.githubusercontent.com/OlivierBinette/b77cc476f34cb6bf706844d395f91e23/raw/cecada69add82dcdb4333e7133ea68d65bc2f972/prettyplot.R"))
```

\small
```{r, fig.width=4, fig.height=3, fig.align="center", eval=FALSE}
f <- function(x) sin(pi*x)^2
unif <- Vectorize(function(x) 1)

x = seq(0,1, length.out = 200)
plot(x, f(x), type="l", col=1, lwd=2, ylab="")
lines(x, unif(x), col=2, lwd=2)

legend("topleft", legend=c("f", "unif"), 
       col=cmap.knitr(c(1,2)), lwd=2, lty=1, cex=0.7)
```
\normalfont


# Lab 5

**Task 1:** Plot $f(x)$ and the Unif(0,1) density. Sample from $f(x)$ using the Unif(0,1) pdf as an enveloping function.

```{r, fig.width=4, fig.height=3, fig.align="center", echo=F}
#source(url("https://gist.githubusercontent.com/OlivierBinette/b77cc476f34cb6bf706844d395f91e23/raw/cecada69add82dcdb4333e7133ea68d65bc2f972/prettyplot.R"))

f <- function(x) sin(pi*x)^2
unif <- Vectorize(function(x) 1)

x = seq(0,1, length.out = 200)
plot(x, f(x), type="l", col=1, lwd=2, ylab="")
lines(x, unif(x), col=2, lwd=2)

legend("topleft", legend=c("f", "unif"), 
       col=cmap.knitr(c(1,2)), lwd=2, lty=1, cex=0.7)
```


# Lab 5

**Task 1:** Plot $f(x)$ and the Unif(0,1) density. Sample from $f(x)$ using the Unif(0,1) pdf as an enveloping function.

Let's implement rejection sampling for a single data point:

\small
```{r}
sample = NULL
while (is.null(sample)) {
  # Step 1
  x = runif(1, min=0, max=1)
  y = runif(1, min=0, max=unif(x))
  
  # Step 2
  if (y < f(x)) sample = x
}

x
```
\normalfont

# Lab 5

Now let's code a more general rejection sampling method.

\small
```{r}
rejection_sampling <- function(f, g, rg) {
  while (TRUE) { # Bad practice; doing this for brevity here.
    # Step 1
    x = rg(1)
    y = runif(1, min=0, max=g(x))
    
    # Step 2
    if (y < f(x)) return(sample)
  }
}

rejection_sampling(f, unif, runif)
```
\normalfont

# Lab 5

**Task 2** Plot a histogram of the points that fall in the acceptance region. Do this for a simulation size of $10^2$ and $10^5$ and report your acceptance ratio. Compare the ratios and histograms.

\small
```{r, fig.width=4, fig.height=3, fig.align="center", eval=FALSE}
k = 10^2

x = runif(k)
y = runif(k)

samples = x[y < f(x)]

hist(samples)
mean(y < f(x))
```
\normalfont

# Lab 5

**Task 2** Plot a histogram of the points that fall in the acceptance region. Do this for a simulation size of $10^2$ and $10^5$ and report your acceptance ratio. Compare the ratios and histograms.

\small
```{r, fig.width=4, fig.height=3, fig.align="center", echo=FALSE}
k = 10^2

x = runif(k)
y = runif(k)

samples = x[y < f(x)]

hist(samples)
mean(y < f(x))
```
\normalfont






# Lab 5

**Task 2** Plot a histogram of the points that fall in the acceptance region. Do this for a simulation size of $10^2$ and $10^5$ and report your acceptance ratio. Compare the ratios and histograms.

\small
```{r, fig.width=4, fig.height=3, fig.align="center", eval=FALSE}
k = 10^6

x = runif(k)
y = runif(k)

samples = x[y < f(x)]

hist(samples)
mean(y < f(x))
```
\normalfont

# Lab 5

**Task 2** Plot a histogram of the points that fall in the acceptance region. Do this for a simulation size of $10^2$ and $10^5$ and report your acceptance ratio. Compare the ratios and histograms.

\small
```{r, fig.width=4, fig.height=3, fig.align="center", echo=FALSE}
k = 10^6

x = runif(k)
y = runif(k)

samples = x[y < f(x)]

hist(samples)
mean(y < f(x))
```
\normalfont

# Lab 5

**Task 3** Repeat Tasks 1 - 2 for Beta(2,2) as an enveloping function.

\small
```{r}
g = function(x) dbeta(x, 2,2)
rg = function(n) rbeta(n, 2,2)
```
\normalfont

\small
```{r, fig.width=4, fig.height=3, fig.align="center", eval=FALSE}
f <- function(x) sin(pi*x)^2
unif <- Vectorize(function(x) 1)

x = seq(0,1, length.out = 200)
plot(x, f(x), type="l", col=1, lwd=2, ylab="", ylim=c(0,1.5))
lines(x, unif(x), col=2, lwd=2)
lines(x, g(x), col=3, lwd=2)

legend("topleft", legend=c("f", "unif", "Beta(2,2)"), 
       col=cmap.knitr(c(1,2,3)), lwd=2, lty=1, cex=0.7)
```
\normalfont


# Lab 5

**Task 3** Repeat Tasks 1 - 2 for Beta(2,2) as an enveloping function.

\small
```{r, fig.width=4.5, fig.height=4, fig.align="center", echo=FALSE}
f <- function(x) sin(pi*x)^2
unif <- Vectorize(function(x) 1)

x = seq(0,1, length.out = 200)
plot(x, f(x), type="l", col=1, lwd=2, ylab="", ylim=c(0,1.5))
lines(x, unif(x), col=2, lwd=2)
lines(x, g(x), col=3, lwd=2)

legend("topleft", legend=c("f", "unif", "Beta(2,2)"), 
       col=cmap.knitr(c(1,2,3)), lwd=2, lty=1, cex=0.6)
```

# Lab 5

**Task 3** Repeat Tasks 1 - 2 for Beta(2,2) as an enveloping function.

\small
```{r}
g = function(x) dbeta(x, 2,2)
rg = function(n) rbeta(n, 2,2)
```

```{r}
rejection_sampling(f, g, rg)
```
\normalfont

# Lab 5

**Task 3** Repeat Tasks 1 - 2 for Beta(2,2) as an enveloping function.

\small
```{r, fig.width=4, fig.height=3, fig.align="center", eval=FALSE}
k = 10^2

x = rg(k)
y = runif(k, min=0, max=g(x))

samples = x[y < f(x)]

hist(samples)
mean(y < f(x))
```
\normalfont

# Lab 5

**Task 3** Repeat Tasks 1 - 2 for Beta(2,2) as an enveloping function.

\small
```{r, fig.width=4, fig.height=3, fig.align="center", echo=FALSE}
k = 10^2

x = rg(k)
y = runif(k, min=0, max=g(x))

samples = x[y < f(x)]

hist(samples)
mean(y < f(x))
```
\normalfont

# Lab 5

**Task 3** Repeat Tasks 1 - 2 for Beta(2,2) as an enveloping function.

\small
```{r, fig.width=4, fig.height=3, fig.align="center", eval=FALSE}
k = 10^6

x = rg(k)
y = runif(k, min=0, max=g(x))

samples = x[y < f(x)]

hist(samples)
mean(y < f(x))
```
\normalfont

# Lab 5

**Task 3** Repeat Tasks 1 - 2 for Beta(2,2) as an enveloping function.

\small
```{r, fig.width=4, fig.height=3, fig.align="center", echo=FALSE}
k = 10^6

x = rg(k)
y = runif(k, min=0, max=g(x))

samples = x[y < f(x)]

hist(samples)
mean(y < f(x))
```
\normalfont
