---
title: "Lab 2 -- Beta-Binomial Distribution"
author: "Rebecca C. Steorts"
date: "January 2018"
output: pdf_document
---

In class, you saw the Binomial-Beta model. We will now use this to solve a very real problem! Suppose I wish to determine whether the probability that a worker will fake an illness is truly 1\%. Your task is to assist me! Tasks 1--3 will be completed in lab and tasks 3--5 should be completed in your weekly homework assignment. You should still upload task 3 even though this will be worked through in lab! 

# Task 1

Let's start by quickly deriving the Beta-Binomial distribution. 

We assume that 
$$X\mid \theta \sim \text{Binomial} (\theta)$$,
$$\theta \sim \text{Beta}(a,b),$$
where $a,b$ are assumed to be known parameters. What is the posterior distribution of $\theta \mid X$?

\begin{align}
p(\theta \mid X) &\propto 
p(X \mid \theta) p(\theta) \\
&\propto \theta^{x} 
(1 - \theta)^{(n-x)} \times \theta^{(a-1)} (1 - \theta)^{(b-1)}\\
&\propto \theta^{x + a -1} (1 - \theta)^{(n-x + b -1)}.
\end{align}

This implies that 
$$\theta \mid X \sim \text{Beta}(x+a,n-x+b).$$



# Task 2

Simulate some data using the \textsf{rbinom} function of size $n = 100$ and probability equal to 1\%. Remember to \textsf{set.seed(123)} so that you can replicate your results.

The data can be simulated as follows:
```{r,echo=TRUE}
# set a seed
set.seed(123)
# create the observed data
obs.data <- rbinom(n = 100, size = 1, prob = 0.01)
# inspect the observed data
head(obs.data)
tail(obs.data)
length(obs.data)
```

# Task 3

Write a function that takes as its inputs that data you simulated (or any data of the same type) and a sequence of $\theta$ values of length 1000 and produces Likelihood values based on the Binomial Likelihood. Plot your sequence and its corresponding Likelihood function.

The likelihood function is given below. Since this is a probability and is only valid over the interval from $[0, 1]$ we generate a sequence over that interval of length 1000.

You have a rough sketch of what you should do for this part of the assignment. Try this out in lab on your own. 

```{r, echo = TRUE}
### Bernoulli LH Function ###
# Input: obs.data, theta
# Output: bernoulli likelihood


### Plot LH for a grid of theta values ###
# Create the grid #
# Store the LH values
# Create the Plot
```



# Task 4 (To be completed for homework)

Write a function that takes as its inputs  prior parameters \textsf{a} and \textsf{b} for the Beta-Bernoulli model and the observed data, and produces the posterior parameters you need for the model. \textbf{Generate and print} the posterior parameters for a non-informative prior i.e. \textsf{(a,b) = (1,1)} and for an informative case \textsf{(a,b) = (3,1)}}.



# Task 5 (To be completed for homework)

Create two plots, one for the informative and one for the non-informative case to show the posterior distribution and superimpose the prior distributions on each along with the likelihood. What do you see? Remember to turn the y-axis ticks off since superimposing may make the scale non-sense.