---
title: "Lab 6: Monte Carlo"
author: "Olivier Binette and Michael Christensen"
date: "Friday September 18, 2020"
fontsize: 11pt
output: 
  beamer_presentation:
    include:
      in_header: preamble.tex
---

```{r, echo=FALSE, include=FALSE}
set.seed(1)
```


# Agenda

1. Review of Monte Carlo and Importance Sampling
2. Lab 6 Tasks 1-3
3. Questions / Office Hours

# 
\section{Review of Monte Carlo and Importance Sampling}


# Review of Monte Carlo and Importance Sampling

**Goal:** Approximate an integral
$$
  I = \int_{\mathcal{X}} h(x) f(x)\, dx = \mathbb{E}_{f}[h(x)]
$$
which is intractable, where $f(x)$ is a probability density function.
\pause

**What's the problem?**
\pause

- Typically $h(x)$ is messy or high-dimensional. We need numerical techniques.
\pause

- In dimension $d=3$ or higher (e.g. $\mathcal{X} = \mathbb{R}^3$), Monte Carlo typically improves upon numerical integration techniques.

# Review of Monte Carlo and Importance Sampling

We want to approximate $I = \int_{\mathcal{X}} h(x) f(x)\, dx = \mathbb{E}_{f}[h(x)]$...

**Monte-Carlo solution:**\pause Sample $X_1, X_2, X_3, \dots, X_n$ from $f$, and estimate $I$ by the empitical average
$$
  \overline h_n = \frac{1}{n}\sum_{i=1}^n h(x_i).
$$
\pause

- The estimate $\overline h_n$ converges almost surely to $I$ as $n \rightarrow \infty$ by the strong law of large numbers.

# Review of Monte Carlo and Importance Sampling

We want to approximate $I = \int_{\mathcal{X}} h(x) f(x)\, dx = \mathbb{E}_{f}[h(x)]$...

**Importance Sampling:**\pause Maybe it's hard to sample from $f$, and instead you'd like to take your samples from a density $g$. So divide and multiply by $g(x)$ to write
$$
  I = \int_{\mathcal{X}} h(x) f(x)\, dx = \int_{\mathcal{X}}h(x)  \frac{f(x)}{g(x)} g(x)\, dx = \mathbb{E}_g\left[h(x)  \frac{f(x)}{g(x)}\right].
$$
\pause

Now use a Monte-Carlo estimate of $I$ with respect to $g$: sample $x_1, x_2, \dots, x_n$ from $g$ and estimate $I$ by
$$
  \hat I = \frac{1}{n}\sum_{i=1}^n h(x_i)\frac{f(x_i)}{g(x_i)}.
$$

#
\section{Lab 6}

# Lab 6

Consider
$$
  I = \int_{-\infty}^\infty e^{-x^4}\,dx.
$$
\pause

1. Find a closed form solution to $I$ and evaluate this.
2. Approximate $I$ using Monte-Carlo.
3. Approximate $I$ using importance sampling

# Lab 6

## Task 1: Find a closed form solution to $I$.

How can we compute $I = \int_{-\infty}^\infty e^{-x^4}\,dx$?
\pause

- Change of variable ($u$-substitution)!
\pause

- First write $I = 2\int_{0}^\infty e^{-x^4}\,dx$.
\pause

- Set $u = x^4$, \pause$x = u^{1/4}$.\pause Then $du = 4x^3 dx$,\pause $dx = \frac{du}{4x^3} = \frac{du}{4 u^{3/4}}$.\pause
\begin{align*}
I &= 2\int_{-\infty}^\infty e^{-x^4}\,dx\\\pause
  &= 2\int_{0}^\infty \frac{e^{-u}}{4u^{3/4}}\,du\\\pause
  &= \frac{1}{2}\int_{0}^\infty u^{-3/4} e^{-u}\,du\\ \pause
  &= \frac{1}{2}\int_{0}^\infty \underbrace{u^{-3/4} e^{-u}}_{\text{Gamma kernel}}\,du
\end{align*}

# Lab 6

## Task 1: Find a closed form solution to $I$.

Remember that the Gamma density is $\frac{b^a}{\Gamma(a)} x^{a-1}e^{-bx}$.\pause

So $\int_0^\infty\frac{b^a}{\Gamma(a)} x^{a-1}e^{-bx}\,dx = 1$
\pause
and $\int_0^\infty x^{a-1}e^{-bx}\,dx = \frac{\Gamma(a)}{b^a}.$
\pause

In our context, we therefore have
\begin{align*}
I &= \frac{1}{2}\int_{0}^\infty \underbrace{u^{-3/4} e^{-u}}_{\text{Gamma kernel}}\,du\\\pause
  &= \frac{1}{2}\int_{0}^\infty u^{1/4 - 1} e^{-u}\,du\\\pause
  &=  \frac{1}{2} \frac{\Gamma(1/4)}{1^{1/4}}\\\pause
  &= \frac{\Gamma(1/4)}{2}\\\pause
  &= 1.813
\end{align*}

# Lab 6

## Task 2: Estimate $I = \int_{-\infty}^\infty e^{-x^4}\,dx$ using Monte-Carlo.
\pause

Beka suggests the substitution $y = \sqrt{2}x^2$.\pause Then
$$
  I = 2^{-5/4}\int_0^\infty \sqrt{\frac{2\pi}{y}} 2\phi(y) \,dy
$$
where $2\phi(y)$ is the density of the normal distribution truncated to $[0,\infty)$.
\pause

So if $Y\sim N(0,1)$, then $Y = |X|$ has density $2\phi$.\pause

**Monte-Carlo algorithm:**\pause

1. Sample $X_1, X_2, \dots, X_n \sim N(0,1)$.\pause
2. Approximate $I$ by 
$$
  \hat I_{MC} = \frac{1}{n}\sum_{i=1}^n 2^{-5/4}\sqrt{\frac{2\pi}{|X_i|}}.
$$
# Lab 6

# Lab 6

## Task 2: Estimate $I = \int_{-\infty}^\infty e^{-x^4}\,dx$ using Monte-Carlo.

Let's implement this:\pause

\small
```{r, echo=TRUE, include=TRUE}
integrand <- function(x) {
  2^{-5/4} * sqrt(2*pi/abs(x))
}

n = 10^6
X = rnorm(n)
values = integrand(X)
mean(values)
sd(values)/sqrt(n)
```
\normalfont


# Lab 6

## Task 3: Estimate $I$ using importance sampling.
\pause

Let's use a *Normal* instrumental distribution:\pause

\begin{align*}
  I &= \int_{-\infty}^\infty e^{-x^4}\,dx\\\pause
    &= \int_{-\infty}^\infty \frac{e^{-x^4}}{\phi(x)}\phi(x)\,dx\pause
\end{align*}
where $\phi$ is the normal density.\pause

**Importance sampling algorithm:**\pause

1. Sample $X_1, X_2, \dots, X_n \sim N(0,1)$.\pause
2. Estimate $I$ by
$$
  \hat I_{IS} = \frac{1}{n}\sum_{i=1}^n \frac{e^{-X_i^4}}{\phi(X_i)}.
$$
\pause

# Lab 6

## Task 3: Estimate $I$ using importance sampling.

Let's implement this:\pause

\small
```{r}
integrand_IS <- function(x) {
  exp(-x^4)/dnorm(x)
}

n = 10^6
X = rnorm(n)
values_IS = integrand_IS(X)
mean(values_IS)
sd(values_IS)/sqrt(n)
```
\normalfont








 








