---
title: "Lab 5"
author: "Andrea Aveni"
date: "2023-02-06"
output: pdf_document
---

# Rejection sampling

**The problem:** We want to sample from a distribution on $\mathbb{R}$ given either:

1. its probability density function $p(\theta)$; or

2. some function $f(\theta) \propto p(\theta)$.

For instance, in a Bayesian framework, we want to sample from the posterior $p(\theta \mid x) \propto p(x \mid \theta) p(\theta)$.

- Rejection sampling is *one* way to address this.

The area under the graph of a function $f$ is the set of points $(x,y)$ such that $0 \leq y \leq f(x)$.

**Fundamental lemma of rejection sampling:** Let $f$ be a positive and integrable function. If $(X,Y)$ is uniformly distributed under the graph of $f$, then the marginal probability density function of $X$ is proportional to $f$.

## Rejection sampling Algorithm

**Input:**

- An integrable function $f \geq 0$ and an enveloppe $g \geq f$ which you can sample from.

**Output:**

- A sample $X$ distributed following the density proportional to $f$.

**Algorithm:**

1. Sample $X \sim g$ and $Y \mid X \sim \text{unif}(0, g(X))$.

2. If $Y < f(X)$, then return $X$. Otherwise go back to step 1.

# Rejection sampling

**Note:**

- The function $g$ is called the *envelop* function, and the corresponding distribution is the *proposal* distribution, or the *instrumental* distribution.

- The function $f$ is called the *target*.

# Lab 5

```{r}
library(ggplot2)
library(tictoc)
```

We want to sample from a random variable with the density
$$f(x) = 2\sin^2(\pi x)\mathbf1_{[0,1]}(x),$$
using rejection sampling. (PS can you verify that this is a density?)

We start considering the proposal $\mathrm{Unif}(0,1)=\mathrm{Beta}(1,1)$, so that its density is $g(x)=\mathbf1_{[0,1]}(x)$.

Before doing anything else we rescale $f$ so that $\alpha f$ is always below the proposal $g$. In order to find the right constant, we need to solve the problem
$$\forall x\in [0,1],\,\alpha f(x)\leq g(x),\,\,\,\,\,\Rightarrow\,\,\,\,\, \alpha\leq  \min_{x\in[0,1]}\frac{g(x)}{f(x)}$$
And since a larger alpha will increase the acceptance rate, we will take
$$\alpha=\min_{x\in[0,1]}\frac{g(x)}{f(x)}$$

```{r}
f=function(x){ return(2*sin(pi*x)^2) }
g1=function(x){ return(dunif(x,0,1)) }
rg1=function(n){ return(rbeta(n,1,1)) }

N=10^5
xx=(1:(N-1))/N
a1=min(g1(xx)/f(xx))

f1=function(x){ return(a1*f(x)) }
```

```{r, fig.align="center"}
ggplot()+
  geom_function(fun=f1,col="blue")+
  geom_function(fun=g1,col="black")+
  xlim(0,1)+theme_bw()
```

We now illustrate the algorithm

```{r}
N=10^5
tic()
Xaccepted=Xrejected=Yaccepted=Yrejected=c()
for(j in 1:N){
  
  x=rg1(1)
  y=runif(1,min=0,max=g1(x))
  
  if(y<f1(x)){
    Xaccepted=c(Xaccepted,x)
    Yaccepted=c(Yaccepted,y)
  }
  else{
    Xrejected=c(Xrejected,x)
    Yrejected=c(Yrejected,y)
  }
  
}
fortime=toc()
```

```{r, fig.align="center"}
ggplot()+
  geom_point(aes(x=Xaccepted,y=Yaccepted),col="blue",size=0.01)+
  geom_point(aes(x=Xrejected,y=Yrejected),col="red",size=0.01)+
  xlab("X")+ylab("Y")+
  theme_bw()
```

We can vectorize this for-loop and speed up computations in the following way

```{r}
tic()
X = rg1(N)
Y = runif(N, min=0, max=g1(X))
acc=(Y<f1(X))
Xaccepted=X[acc]
Yaccepted=Y[acc]
Xrejected=X[!acc]
Yrejected=Y[!acc]
vectorizedtime=toc()
```

```{r, fig.align="center"}
ggplot()+
  geom_point(aes(x=Xaccepted,y=Yaccepted),col="blue",size=0.1)+
  geom_point(aes(x=Xrejected,y=Yrejected),col="red",size=0.1)+
  xlab("X")+ylab("Y")+
  theme_bw()
```

As you can see the result is the same but, by vectorizing, we reduced the time by a huge factor of
```{r}
as.numeric((fortime$toc-fortime$tic)/(vectorizedtime$toc-vectorizedtime$tic))
```

We now provide a general function for rejection sampling

```{r}
rej_sampl=function(N,f,g,rg){
  X = rg(N)
  Y = runif(N, min=0, max=g(X))
  acc=(Y<f(X))
  Xaccepted=X[acc]
  return(Xaccepted)
}

head(rej_sampl(1000,f1,g1,rg1))
```


We plot a histogram of the points that fall in the acceptance region for different simulation sizes and we report our acceptance ratio. We compare the ratios and histograms.

```{r, fig.align="center"}
N=10^2
S=rej_sampl(N,f1,g1,rg1)

length(S)
ggplot()+
  geom_density(aes(x=S))+
  geom_function(fun=f,col="blue")+
  xlim(0,1)+xlab("X")+ylab("Density")+
  theme_bw()
```

```{r, fig.align="center"}
N=10^5
S=rej_sampl(N,f1,g1,rg1)

length(S)
ggplot()+
  geom_density(aes(x=S))+
  geom_function(fun=f,col="blue")+
  xlim(0,1)+xlab("X")+ylab("Density")+
  theme_bw()
```

Thus in this case the rejection sampling is $1/2$. And this is not surprisingly, since the acceptance rate is equal to the area under the rescaled $f$. Thus
$$\textrm{rejection rate}=\int_0^1\alpha f(x)\mathrm{d}x=\alpha$$
And, indeed, using the uniform proposal we got $\alpha=1/2$.

Now we consider a different proposal that will lead to better rejection rates. We consider
$\mathrm{Beta}(3,3)$ density.

```{r}
g2=function(x){ return(dbeta(x,3,3)) }
rg2=function(n){ return(rbeta(n,3,3)) }

N=10^5
xx=(1:(N-1))/N
a2=min(g2(xx)/f(xx))

f2=function(x){ return(a2*f(x)) }
```

In this case the acceptance rate that we get is much higher (15/16), let's visualize this

```{r, fig.align="center"}
ggplot()+
  geom_function(fun=f2,col="blue")+
  geom_function(fun=g2,col="black")+
  xlab("X")+ylab("Density")+xlim(0,1)+
  theme_bw()
```
And now let's proceed with the rejection sampling

```{r}
tic()
X = rg2(N)
Y = runif(N, min=0, max=g2(X))
acc=(Y<f2(X))
Xaccepted=X[acc]
Yaccepted=Y[acc]
Xrejected=X[!acc]
Yrejected=Y[!acc]
vectorizedtime=toc()

sum(acc)/N
```

```{r, fig.align="center"}
ggplot()+
  geom_point(aes(x=Xaccepted,y=Yaccepted),col="blue",size=0.01)+
  geom_point(aes(x=Xrejected,y=Yrejected),col="red",size=0.01)+
  xlab("X")+ylab("Y")+
  theme_bw()
```

Plot a histogram of the points that fall in the acceptance region. Do this for a simulation size of $10^2$ and $10^5$ and report your acceptance ratio. Compare the ratios and histograms.

```{r, fig.align="center"}
N=10^2
S=rej_sampl(N,f2,g2,rg2)

length(S)/N
ggplot()+
  geom_density(aes(x=S))+
  geom_function(fun=f,col="blue")+
  xlim(0,1)+xlab("X")+ylab("Density")+
  theme_bw()
```

```{r, fig.align="center"}
N=10^5
S=rej_sampl(N,f2,g2,rg2)

length(S)/N
ggplot()+
  geom_density(aes(x=S))+
  geom_function(fun=f,col="blue")+
  xlim(0,1)+xlab("X")+ylab("Density")+
  theme_bw()
```


