---
title: 'Introduction to R, Part III'
author: "Rebecca C. Steorts, STA 360"
output: pdf_document
---

Agenda
=======
- Example of housing in PA
- Review of linear models
- Using variables and names versus hard coding


Example: Price of houses in PA
===
Census data for California and Pennsylvania on housing prices,
by Census "tract"

```{r}
# read in data
calif_penn <-
  read.csv("http://www2.stat.duke.edu/~rcs46/modern_bayes17/data/calif_penn_2011.csv")
# inspect the variables associated with this dataset
names(calif_penn)
# STATEFP is the FIPS code, where there is one for each state. 42 belongs to PA.
# 6 corresponds to CA. 
# https://en.wikipedia.org/wiki/Federal_Information_Processing_Standard_state_code#FIPS_state_codes
penn <- calif_penn[calif_penn[,"STATEFP"]==42,]
# fitting a simple linear model 
coefficients(lm(Median_house_value ~ Median_household_income, data=penn))
summary(lm(Median_house_value ~ Median_household_income, data=penn))
```


Goal: fit a simple linear model, and predict the median house price (y) from median household income (x). Before doing this, 
let's investigate the census tracts that correspond to Allegheny county (24---425).

Tract 24 has a median income of \$14,719; actual median house value is \$34,100 --- is that above or below the observed median?
```{r}
34100 < -26206.564 + 3.651*14719
```

Tract 25 has income \$48,102 and house price \$155,900
```{r}
155900 < -26206.564 + 3.651*48102
```

What about tract 26?

We _could_ just keep plugging in numbers like this, but that's

- boring and repetitive
- error-prone
- confusing (what _are_ these numbers?)

Using variables and names
===
```{r}
penn.coefs <- coefficients(lm(Median_house_value ~ Median_household_income, data=penn))
penn.coefs
```

```{r}
allegheny.rows <- 24:425
allegheny.medinc <- penn[allegheny.rows,"Median_household_income"]
allegheny.values <- penn[allegheny.rows,"Median_house_value"]
allegheny.fitted <- penn.coefs["(Intercept)"]+
  penn.coefs["Median_household_income"]*allegheny.medinc
```

Actual median house values versus Predicted Median House Values 
===
```{r}
plot(x=allegheny.fitted, y=allegheny.values,
     xlab="Model-predicted median house values",
     ylab="Actual median house values",
     xlim=c(0,5e5),ylim=c(0,5e5))
abline(a=0,b=1,col="grey")
```

Summary
===

- We have reviewed simple linear models. 
- We used variable and naming schemes.
- We reviewed how to plot. 
- We have looked at a real application from the Census in the state of PA, where we avoided
hard coding for easy automation. 

