---
title: 'Introduction to R, Part IV'
author: "Rebecca C. Steorts, STA 360"
output: pdf_document
---

Agenda
=======
- Matrices
- Matrix Operations
- Examples 


Simple example: resource allocation ("mathematical programming")
====
Factory makes cars and trucks, using labor and steel

- a car takes 40 hours of labor and 1 ton of steel
- a truck takes 60 hours and 3 tons of steel
- resources: 1600 hours of labor and 70 tons of steel each week

Matrices
===
In R, a matrix is a specialization of a 2D array

```{r}
factory <- matrix(c(40,1,60,3),nrow=2)
is.array(factory)
is.matrix(factory)
```
could also specify `ncol`, and/or `byrow=TRUE` to fill by rows.

Element-wise operations proceed as usual (e.g., `factory/5`)

Matrix multiplication
===
Gets a special operator

```{r}
six.sevens <- matrix(rep(7,6),ncol=3)
six.sevens
factory %*% six.sevens # [2x2] * [2x3]
```

Exercise: What if you try `six.sevens %*% factory`?

Multiplying matrices and vectors
===
Numeric vectors can act like proper vectors:
```{r}
output <- c(10,20)
factory %*% output
output %*% factory
```
<small>R silently casts the vector as either a row or a column matrix</small>

Matrix operators
===

Transpose:
```{r}
t(factory)
```

Determinant:
```{r}
det(factory)
```

The diagonal
===
The `diag()` function can extract the diagonal entries of a matrix:
```{r}
diag(factory)
```

Creating a diagonal or identity matrix
===
```{r}
diag(c(3,4))
diag(2)
diag(10)
```

Inverting a matrix
===
```{r}
solve(factory)
factory %*% solve(factory)
```

Why's it called "solve"" anyway?
===
Solving the linear system $\mathbf{A}\vec{x} = \vec{b}$ for $\vec{x}$:

```{r}
# this is acting like b in our linear system
# factory is behaving like A
# goal is to solve for x
available <- c(1600,70)
solve(factory,available)
factory %*% solve(factory,available)
```

Names in matrices
===
We can name either rows or columns or both, with `rownames()` and `colnames()`

These are character vectors

We use the same function to get and to set their respective values

Names are useful since they help us keep track of what
we are working with

Example
===
```{r}
rownames(factory) <- c("labor","steel")
colnames(factory) <- c("cars","trucks")
factory
available <- c(1600,70)
names(available) <- c("labor","steel")
```

Example (Continued)
===
```{r}
output <- c(20,10)
names(output) <- c("cars","trucks")
factory %*% output
factory %*% output[colnames(factory)]
all(factory %*% output[colnames(factory)] <= available[rownames(factory)])
```

Summaries
===
Take the mean: `rowMeans()`, `colMeans()`: input is matrix,
output is vector.  Also `rowSums()`, etc.

`summary()`: vector-style summary of column

```{r}
colMeans(factory)
summary(factory)
```

Apply function
===
`apply()`, takes 3 arguments: the array or matrix, then 1 for rows and 2 for columns, then name of the function to apply to each

```{r}
rowMeans(factory)
apply(factory,1,mean)
```

<small>What would `apply(factory,1,sd)` do?</small>
