---
title: 'Introduction to R, Part V'
author: "Rebecca C. Steorts, STA 360"
output: pdf_document
---

Agenda
=======
- Lists
- Dataframes


Lists
====
Sequence of values, _not_ necessarily all of the same type

```{r}
my.distribution <- list("exponential",7,FALSE)
my.distribution
```

Most of what you can do with vectors you can also do with lists

Accessing pieces of lists
===
Can use `[ ]` as with vectors
or use `[[ ]]`, but only with a single index
`[[ ]]` drops names and structures, `[ ]` does not
```{r}
is.character(my.distribution)
is.character(my.distribution[[1]])
my.distribution[[2]]^2
```
<small>What happens if you try `my.distribution[2]^2`?</small>
<small>What happens if you try `[[ ]]` on a vector?</small>

Expanding and contracting lists
===
Add to lists with `c()` (also works with vectors):
```{r}
my.distribution <- c(my.distribution,7)
my.distribution
```


Chop off the end of a list by setting the length to something smaller (also works with vectors):
```{r}
length(my.distribution)
length(my.distribution) <- 3
my.distribution
```

Naming list elements
===
We can name some or all of the elements of a list
```{r}
names(my.distribution) <- c("family","mean","is.symmetric")
my.distribution
my.distribution[["family"]]
my.distribution["family"]
```


Lists have a special short-cut way of using names, `$` (which removes names and structures):
```{r}
my.distribution[["family"]]
my.distribution$family
```

Names in lists (cont'd.)
===
Creating a list with names:
```{r}
another.distribution <- list(family="gaussian",mean=7,sd=1,is.symmetric=TRUE)
```

Adding named elements:
```{r}
my.distribution$was.estimated <- FALSE
my.distribution[["last.updated"]] <- "2011-08-30"
```

Removing a named list element, by assigning it the value `NULL`:
```{r}
my.distribution$was.estimated <- NULL
```

Key-Value pairs
===
Lists give us a way to store and look up data by _name_, rather than by _position_

A really useful programming concept with many names: **key-value pairs**, **dictionaries**, **associative arrays**, **hashes**

If all our distributions have components named `family`, we can look that up by name, without caring where it is in the list

Dataframes
===
Dataframe = the classic data table, $n$ rows for cases, $p$ columns for variables

Lots of the really-statistical parts of R presume data frames
<small>`penn` from last time was really a dataframe</small>

Not just a matrix because *columns can have different types*

Many matrix functions also work for dataframes (`rowSums()`, `summary()`, `apply()`)

<small>but no matrix multiplying dataframes, even if all columns are numeric</small>


```{r}
a.matrix <- matrix(c(35,8,10,4),nrow=2)
colnames(a.matrix) <- c("v1","v2")
a.matrix
a.matrix[,"v1"]  # Try a.matrix$v1 and see what happens
```


```{r}
a.data.frame <- data.frame(a.matrix,logicals=c(TRUE,FALSE))
a.data.frame
a.data.frame$v1
a.data.frame[,"v1"]
a.data.frame[1,]
colMeans(a.data.frame)
```

Adding rows and columns
===
We can add rows or columns to an array or data-frame with `rbind()` and `cbind()`, but be careful about forced type conversions
```{r}
rbind(a.data.frame,list(v1=-3,v2=-5,logicals=TRUE))
# Take care here as the command below is changing the dataframe! 
rbind(a.data.frame,c(3,4,6))
```

Back to the Factory Example 
===
Recall from the last module, we had an example with a factory, where
a factory makes cars and trucks, using labor and steel

- a car takes 40 hours of labor and 1 ton of steel
- a truck takes 60 hours and 3 tons of steel
- resources: 1600 hours of labor and 70 tons of steel each week

```{r}
factory <- matrix(c(40,1,60,3),nrow=2)
output <- c(10,20)
available <- c(1600,70)
```

Structures of Structures
===
So far, every list element has been a single data value

List elements can be other data structures, e.g., vectors and matrices:
```{r}
plan <- list(factory=factory, available=available, output=output)
plan$output
```

Structures of Structures (cont'd.)
===
List elements can even be other lists
which may contain other data structures
including other lists
which may contain other data structures...

This **recursion** lets us build arbitrarily complicated data structures from the basic ones

Most complicated objects are (usually) lists of data structures



Take-Aways
===
- Write programs by composing functions to manipulate data
- The basic data types let us represent Booleans, numbers, and characters
- Data structures let us group related values together
- Vectors let us group values of the same type
- Use variable assignment and name components of structures to make data more meaningful
- Matrices act like you'd hope they would
- Lists let us combine different types of data
- Dataframes are hybrids of matrices and lists, for classic tabular data
