---
title: 'Introduction to R, Part II'
author: "Rebecca C. Steorts, STA 360"
output: pdf_document
---

Agenda
=======
- Data Structures
- Vectors

Data Structure
===

A **data structure** is group related data values that are stored in one object. 

An example of a **data structure** is a **vector**, which is a sequence of values (all of the same type). 

Vectors
===

```{r}
x <- c(7, 8, 10, 45)
```

`c()` function returns a vector containing all its arguments in order

`x[1]` is the first element, `x[3]` is the 3rd element
`x[-3]` is a vector containing all but the 3rd element

Exercise: What does x[-c(2:3)] return?

`vector(length=6)` returns an empty vector of length 6; helpful for filling things up later


Exercise on Vectors
===
```{r}
weekly.hours <- vector(length=5)
weekly.hours[5] <- 8
weekly.hours
```


Operators apply to vectors "pairwise" or "elementwise":
```{r}
y <- c(-7, -8, -10, -45)
x+y
x*y
```


Can also do pairwise comparisons:
```{r}
x > 9
```
Note: returns Boolean vector




Boolean operators work elementwise:
```{r}
(x > 9) & (x < 20)
```

Functions on vectors
===
Many functions take vectors as arguments:

- `mean()`, `median()`, `sd()`, `var()`, `max()`, `min()`, `length()`, `sum()`: return single numbers
- `sort()` returns a new vector
- `hist()` takes a vector of numbers and produces a histogram, a highly structured object, with the side-effect of making a plot
- Similarly `ecdf()` produces a cumulative-density-function object
- `summary()` gives a five-number summary of numerical vectors
- `any()` and `all()` are useful on Boolean vectors

Addressing vectors
===

Vector of indices:
```{r}
x[c(2,4)]
```

Vector of negative indices
```{r}
x[c(-1,-3)]
```
<small>(why that, and not  `7 10`?)</small>


Boolean vector:
```{r}
x[x>9]
y[x>9]
```


`which()` turns a Boolean vector in vector of TRUE indices:
```{r}
x
y
places <- which(x > 9)
places
y[places]
```

Named components
===

You can give names to elements or components of vectors
```{r}
names(x) <- c("v1","v2","v3","fred")
names(x)
x[c("fred","v1")]
```
note the labels is what R prints; not actually part of the value

`names(x)` is just another vector (of characters):
```{r}
names(y) <- names(x)
sort(names(x))
which(names(x)=="fred")
```