---
title: "Lab 10: Bayesian linear regression"
author: "Olivier Binette"
date: "Friday October 23, 2020"
fontsize: 11pt
output: 
  beamer_presentation:
    include:
      in_header: preamble.tex
---

```{r, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=4.5, fig.height=3.5, fig.align="center")
set.seed(1)
```

# Agenda

- Announcements
- Review of Bayesian linear regression
- Lab 10 

#
\section{Announcements}

# Announcements

- Very stressful week and we're thinking about you.
- We're trying to give the best learning experience despite everything.
  - Feedback is appreciated (box is open until November 18): \url{https://app.suggestionox.com/r/OlivierSuggestionBox}
- Let your TAs know if you are facing any issue.
  - I do not grade and I can bring things up anonymously.


#
\section{Review of Bayesian linear regression}

# Review of Bayesian linear regression

**Model:**

$$
  Y \mid X, \beta, \sigma^2 \sim MVN(X \beta, \sigma^2 I_n)
$$
**Prior (known variance):**
$$
  \beta \sim MVN(\beta_0, \Sigma_0)
$$

# Review of Bayesian linear regression

**Posterior distribution:**

$$
  \beta \mid Y, X, \sigma^2 \sim MVN(\beta_n, \Sigma_n)
$$
where

$$
  \beta_n = \left(\Sigma_0^{-1} + X^T X / \sigma^{2} \right)^{-1}\left( \Sigma_{0}^{-1} \beta_0 + X^T Y /\sigma^2 \right),
$$
$$
  \Sigma_n = \left(\Sigma_0^{-1} + X^T X/\sigma^2\right)^{-1}.
$$

#
\section{Lab 10 (linear regression)}


# Lab 10

Exercice 9.1 in Hoff's book.

\begin{center}
  \url{https://github.com/resteorts/modern-bayes/blob/master/labs/10-linear-regression/11-linear-regression_v2.pdf}
\end{center}

# Lab 10

**The problem:** 

- Data on four swimmers describing the evolution of their lap time up until now.

- We want to (1) predict their lap time two weeks from now and (2) predict who will be the fastest swimmer.

# Lab 10

**More precisely:** 

For each swimmer, we have lap times $Y_i = (Y_{i,1}, Y_{i,2}, \dots, Y_{i,6})$ associated with a week $x_i\in \{1,3,5,7,9,11\}$.

**The approach:**

We'll independently fit linear models for each swimmer.


# Lab 10

Let's first load the data.

\small
```{r, message=FALSE, warning=FALSE}
data = read.table(url("https://raw.githubusercontent.com/resteorts/modern-bayes/master/labs/10-linear-regression/data/swim.dat"), header=FALSE)
weeks = c(1,3,5,7,9,11)
```
\normalfont


# Lab 10

And let's take a look.

\small
```{r, eval=FALSE}
plot(weeks, data[1,], ylim=range(data),
     type="l", col=1, ylab="time")
for (i in 2:4) {
  lines(weeks, data[i,], col=i)
}
legend("topright", legend=paste("Swimmer", 1:4),
       lty=1, col=1:4, cex=0.45)
```
\normalfont

# Lab 10

\small
```{r, echo=FALSE}
week = 1:6

plot(week, data[1,], ylim=range(data),
     type="l", col=1, ylab="time")
for (i in 2:4) {
  lines(week, data[i,], col=i)
}
legend("topright", legend=paste("Swimmer", 1:4),
       lty=1, col=1:4, cex=0.45)
```
\normalfont

# Lab 10

## Task 1

We will fit a separate linear regression model for each swimmer, with swimming time as the response and week as the explanatory variable. Let $Y_{i}\in \mathbb{R}^{6}$ be the 6 recorded times for swimmer $i=1,2,3,4.$ Let
$$
X_i =
\left(\begin{matrix}
    1 & 1  \\
    1 & 3 \\ 
    ... \\
    1 & 9\\
    1 & 11
\end{matrix}\right)
$$
be the design matrix for swimmer $i=1,2,3,4.$ 

# Lab 10

## Task 1

Then we use the following linear regression model: 
\begin{align*}
    Y_i \mid \beta_i, \tau_i &\sim \mathcal{N}_6\left(X\beta_i, \tau_i^{-1}\mathcal{I}_6\right) \\
    \beta_i &\sim \mathcal{N}_2\left(\beta_0, \Sigma_0\right) \\
    \tau_i &\sim \text{Gamma}(a,b).
\end{align*}
Derive full conditionals for $\beta_i$ and $\tau_i.$ Assume that $\beta_0, \Sigma_0, a, b$ are known. 

# Lab 10

## Solution

1. Using the the result on slide 7, derive the full conditional $\beta_i \mid Y_i, X, \sigma^2$.

2. Derive the full conditional $\tau_i \mid Y_i, X, \beta_i$:
  - First write down the likelihood.
  - Next write the joint posterior (up to a normalizing constant).
  - Finally derive the full conditional.

# Lab 10

## Task 2

Complete the prior specification by choosing $a,b,\beta_0,$ and $\Sigma_0.$ Let your choices be informed by the fact that times for this age group tend to be between 22 and 24 seconds. 

# Lab 10

## Task 3

Code a Gibbs sampler to fit each of the models. For each swimmer $i,$ obtain draws from the posterior predictive distribution for $y_i^{*},$ the time of swimmer $i$ if they were to swim two weeks from the last recorded time.

# Lab 10

## Task 4

The coach has to decide which swimmer should compete in a meet two weeks from the last recorded time. Using the posterior predictive distributions, compute $\text{Pr}\left\{y_i^{*}=\text{max}\left(y_1^{*},y_2^{*},y_3^{*},y_4^{*}\right)\right\}$ for each swimmer $i$ and use these probabilities to make a recommendation to the coach. 







